---
title: "dplyr-Style Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr-Style Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes glyexp's dplyr-style functions for synchronized data manipulation.
[SummarizedExperiment](https://bioconductor.org/packages//release/bioc/html/SummarizedExperiment.html)
provides basic ways to inspect and manipulate you SE objects.
[dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) and
[tidyr](https://cran.r-project.org/web/packages/tidyr/index.html)
provides various verb functions to manipulate tabels.
`glyexp` tries to bridge up the two world, just like
[tidySummarizedExperiment](https://bioconductor.posit.co/packages/release/bioc/html/tidySummarizedExperiment.html),
but in a different way.
While `tidySummarizedExperiment` re-use `dplyr` functions by regarding the
`SummarizedExperiment` object as a long-format tibble,
`glyexp` splits each `dplyr` verb into two: one for the row data, the other for the column data.

> **Note:** This vignette assumes you are familiar with both `SummarizedExperiment` and `dplyr`.

> **Another Note:** `glyexp` provides two subclasses of `SummarizedExperiment`: `GlycomicSE` and `GlycoproteomicSE`.
All functions introduced in this vignette work with all three of them.

```{r setup, include=FALSE}
library(glyexp)
library(dplyr)
library(tibble)
library(stringr)
library(S4Vectors)
library(SummarizedExperiment)
library(conflicted)

conflicts_prefer(glyexp::select_var)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
```

First, let's create a `SummarizedExperiment` object.

```{r}
expr_mat <- matrix(1:24, nrow = 4)
rownames(expr_mat) <- c("V1", "V2", "V3", "V4")
colnames(expr_mat) <- c("S1", "S2", "S3", "S4", "S5", "S6")
col_data <- DataFrame(
  group = c("A", "A", "A", "B", "B", "B"),
  batch = c(1, 2, 1, 2, 1, 2),
  row.names = colnames(expr_mat)
)
row_data <- DataFrame(
  protein = c("PRO1", "PRO2", "PRO3", "PRO3"),
  peptide = c("PEP1", "PEP2", "PEP3", "PEP4"),
  glycan_composition = c("H5N2", "H5N2", "H3N2", "H3N2"),
  row.names = rownames(expr_mat)
)
toy_exp <- SummarizedExperiment(
  assays = SimpleList(data = expr_mat),
  rowData = row_data,
  colData = col_data
)
print(toy_exp)
```

## Two Flavors: `_obs()` and `_var()`

Every dplyr-style function in glyexp comes in two variants:

- **`_obs()` functions**: Work on column data (observation metadata) in `SummarizedExperiment` objects
- **`_var()` functions**: Work on row data (variable metadata) in `SummarizedExperiment` objects

Both variants automatically update the expression matrix to maintain synchronization.
The mental model should be:
"I want to update the row data, so I should use `xxx_var()` functions like regular `dplyr` functions,
then the expression matrix will update itself."

Filtering is the most common operation.
Say you want to focus only on group "A" samples:

```{r}
# the SummarizedExperiment way
filtered_exp1 <- toy_exp[, toy_exp$group == "A"]

# the glyexp way
filtered_exp2 <- filter_obs(toy_exp, group == "A")

# same result
identical(filtered_exp1, filtered_exp2)
```

In this simple example, there is no clear benefits of using the glyexp way.
The true benefits come with chaining, when you have more than one operations to do:

```{r}
# the SummarizedExperiment way
filtered_exp3 <- toy_exp
colData(filtered_exp3)$batch <- factor(colData(filtered_exp3)$batch)
colData(filtered_exp3)$group <- factor(colData(filtered_exp3)$group)
filtered_exp3 <- filtered_exp3[
  rowData(filtered_exp3)$glycan_composition %in% c("H5N2", "N3N2"),
  filtered_exp3$group == "A"
]

# the glyexp way
filtered_exp4 <- toy_exp |>
  mutate_obs(group = factor(group), batch = factor(batch)) |>
  filter_obs(group == "A", .drop_levels = FALSE) |>
  filter_var(glycan_composition %in% c("H5N2", "N3N2"))

# the same result
identical(filtered_exp3, filtered_exp4)
```

Now the merits of `glyexp` is clear:
You can use `dplyr`-style data manipulation directly on `SummarizedExperiment` objects,
no matter how long your operation pipeline is.

## Virtual Index Columns

In a `SummarizedExperiment`, `rowData` and `colData` are `DataFrame` objects.
`DataFrame`s can have row names, which can be accessed by `rownames()`.

```{r}
rownames(colData(toy_exp))
rownames(rowData(toy_exp))
```

These row names match the row or column names of the assay (expression matrix):

```{r}
rownames(assay(toy_exp))
colnames(assay(toy_exp))
```

In fact, you can get these names by directly calling `rownames()` or `colnames()`
on the `SummarizedExperiment` object:

```{r}
rownames(toy_exp)
colnames(toy_exp)
```

`dplyr` functions do not have supports to row names operations,
because they were designed to be used on `tibble`s,
which do not have row names.

In `glyexp`, however, we provide a unique techqnue to manipulate the row names
with the `dplyr`-style functions: the **virtual index columns**.

The row names of the row and column data can be referred to as a plain
`.variable` or `.sample` column, respectively.

For example, you can update the row and column names of the `SummarizedExperiment` by:

```{r}
toy_exp |>
  mutate_obs(.sample = str_replace(.sample, "S", "Sample_")) |>
  mutate_var(.variable = paste(protein, peptide, glycan_composition, sep = "-"))
```

To understand it better, you can regard the row and column data as:

```{r}
rownames_to_column(as.data.frame(rowData(toy_exp)), ".variable")
rownames_to_column(as.data.frame(colData(toy_exp)), ".sample")
```

when you use the `dplyr`-style functions.

## Complete Function Reference

glyexp provides dplyr-style equivalents for common data manipulation functions.
Each function comes in both `_obs()` and `_var()` variants, and all automatically maintain matrix synchronization.

### Core Data Manipulation Functions

| Standard dplyr | Sample Operations | Variable Operations | Description |
|:---|:---|:---|:---|
| `filter()` | `filter_obs()` | `filter_var()` | Subset with sync |
| `select()` | `select_obs()` | `select_var()` | Choose with protection |
| `arrange()` | `arrange_obs()` | `arrange_var()` | Sort with order |
| `mutate()` | `mutate_obs()` | `mutate_var()` | Create with consistency |
| `rename()` | `rename_obs()` | `rename_var()` | Rename with safety |

### Advanced Slicing Functions

| Standard dplyr | Sample Operations | Variable Operations | Description |
|:---|:---|:---|:---|
| `slice()` | `slice_obs()` | `slice_var()` | Position-based selection |
| `slice_head()` | `slice_head_obs()` | `slice_head_var()` | Top n with sync |
| `slice_tail()` | `slice_tail_obs()` | `slice_tail_var()` | Bottom n with sync |
| `slice_sample()` | `slice_sample_obs()` | `slice_sample_var()` | Random with consistency |
| `slice_max()` | `slice_max_obs()` | `slice_max_var()` | Highest values with order |
| `slice_min()` | `slice_min_obs()` | `slice_min_var()` | Lowest values with order |

### Joining Functions

| Standard dplyr | Sample Operations | Variable Operations | Description |
|:---|:---|:---|:---|
| `left_join()` | `left_join_obs()` | `left_join_var()` | Add new columns from another table (left join) |
| `inner_join()` | `inner_join_obs()` | `inner_join_var()` | Add new columns from another table (inner join) |
| `semi_join()` | `semi_join_obs()` | `semi_join_var()` | Filter rows from another table (semi join) |
| `anti_join()` | `anti_join_obs()` | `anti_join_var()` | Filter rows from another table (anti join) |

## Function-by-Function Examples

### Selection

```{r}
# Select specific columns from sample info
selected_exp <- select_obs(toy_exp, group)
colData(selected_exp)
```

```{r}
# Select columns from variable info (notice the index protection!)
var_selected_exp <- select_var(toy_exp, glycan_composition)
rowData(var_selected_exp)
```

Use `dplyr`-style helpers like `starts_with()`, `ends_with()`, and `contains()`:

```{r}
# Select columns starting with "glycan"
helper_exp <- select_var(toy_exp, starts_with("glycan"))
rowData(helper_exp)
```

### Arrangement

```{r}
# Arrange samples by batch and group
arranged_exp <- arrange_obs(toy_exp, batch, group)
rowData(arranged_exp)
```

Check how the expression matrix columns rearranged to match:

```{r}
# Expression matrix columns follow the new sample order
assay(arranged_exp)
```

### Mutation

```{r}
# Add a new calculated column to sample info
mutated_exp <- mutate_obs(
  toy_exp,
  group_batch = paste(group, batch, sep = "_")
)
colData(mutated_exp)
```

```{r}
# Extract number of Hex and HexNAc
complex_exp <- mutate_var(
  toy_exp,
  nH = as.integer(str_extract(glycan_composition, "H(\\d+)", 1)),
  nN = as.integer(str_extract(glycan_composition, "N(\\d+)", 1)),
)
rowData(complex_exp)
```

### Slicing

```{r}
# Take the first 2 samples
head_exp <- slice_head_obs(toy_exp, n = 2)
colData(head_exp)
```

```{r}
# Expression matrix automatically adjusts
assay(head_exp)
```

```{r}
# Sample randomly from variables
set.seed(123)  # For reproducibility
random_exp <- slice_sample_var(toy_exp, n = 3)
rowData(random_exp)
```

### Renaming

```{r}
# Rename columns in sample info
renamed_exp <- rename_obs(toy_exp, experimental_group = group)
colData(renamed_exp)
```

The index column "sample" remains protected, but everything else can be renamed freely.

### Joining

These functions can be useful if you have additional information stored in a separate tibble,
and you want to add it to your data container.

```{r}
# Join sample info with variable info
more_sample_info <- tibble::tibble(
  .sample = c("S2", "S3", "S4", "S5", "S6"),
  age = c(21, 22, 23, 24, 25),
  gender = c("F", "M", "F", "M", "F")
)
joined_exp <- left_join_obs(toy_exp, more_sample_info, by = ".sample")
colData(joined_exp)
```

You might have noticed that we don't have alternatives for `dplyr::right_join()` and `dplyr::full_join()`.
This is because by design joining functions in `glyexp` should only be used to add new information to your data container.
However, `right_join()` and `full_join()` will add more observations to the resulting tibbles,
which is not suitable for `SummarizedExperiment` objects.

For the same reason, the `relationship` parameter is fixed to "many-to-one" for all joining functions in `glyexp`.
You probably don't need to know this, but if you do,
check out the documentation of `dplyr::left_join()` for more details.

## When dplyr-Style Functions Cannot Help

Sometimes you need functionality beyond what glyexp's dplyr-style functions provide.
Extract the component and use any function you want.

### Why Doesn't glyexp Implement All dplyr Functions?

glyexp only implements functions that preserve the synchronized multi-table structure of `SummarizedExperiment` objects.

Functions like `count()`, `distinct()`, `summarise()`, and `pull()` return aggregated results
that break the original data relationships.
For these operations, extract the relevant component and use standard dplyr functions:

```{r}
# For complex aggregations
toy_exp |>
  colData() |>
  as_tibble() |>
  count(group)
```

```{r}
# For distinct values
toy_exp |>
  rowData() |>
  as_tibble() |>
  distinct(protein) |>
  pull(protein)
```

```{r}
# Or just
unique(rowData(toy_exp)$protein)
```

## Common Pitfalls and How to Avoid Them

### Pitfall 1: Using glyexp Functions on Non-SE Objects

This won't work:
```{r error=TRUE}
regular_tibble <- tibble(group = c("A", "B"), value = c(1, 2))
filter_obs(regular_tibble, group == "A")
```

### Pitfall 2: Forgetting the Synchronization

Don't do this:
```{r eval=FALSE}
# This will not update `toy_exp` at all!
sample_info <- colData(toy_exp) |> as_tibble()
filtered_samples <- filter(sample_info, group == "A")
```

Do this instead:
```{r}
filtered_exp <- filter_obs(toy_exp, group == "A")
```

### Pitfall 3: Trying to Remove Index Columns

This won't work as expected:
```{r error=TRUE}
select_obs(toy_exp, -.sample)
```

Embrace the protection:
```{r}
clean_exp <- select_obs(toy_exp, group)
colData(clean_exp)
```

### Pitfall 4: Mismatched Operations

Don't mix operations inappropriately:
```{r eval=FALSE}
# `glycan_cimposition` is a column in the row data
# `arrange_obs()` works for column data
arrange_obs(toy_exp, glycan_composition)
```

Use the right function for the right data:
```{r}
arranged_by_composition <- arrange_var(toy_exp, glycan_composition)
```

## Performance Considerations

glyexp's dplyr-style functions are designed to be fast, safe, and consistent.

For large datasets, consider:

- Filtering early in your pipeline to reduce data size
- Using `select_obs()` and `select_var()` to keep only needed columns
- Chaining operations efficiently to minimize intermediate copies

```{r}
# Efficient pipeline: filter first, then manipulate
efficient_pipeline <- toy_exp |>
  filter_obs(group == "A") |>          # Reduce samples early
  filter_var(protein == "PRO1") |>     # Reduce variables early
  select_obs(group) |>                 # Keep only needed sample columns
  select_var(glycan_composition)       # Keep only needed variable columns
```

## Philosophy Behind the Design

glyexp's dplyr-style functions embody a simple philosophy:

**"Think about your metadata, and let the data follow."**

This design means:

1. **Mental Model Alignment**: Think in terms of samples and variables, not matrix indices
2. **Error Prevention**: Automatic synchronization prevents common data analysis mistakes
3. **Familiar Syntax**: If you know dplyr, you already know most of glyexp
4. **Composability**: Functions chain together naturally for complex analyses
