---
title: "Experiment Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experiment Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the world of glycoverse experiments! This vignette will walk you through 
the four distinct experiment types in the `glyexp` package: "glycomics", 
"glycoproteomics", "traitomics", and "traitproteomics". Think of these as 
different flavors of glycobiology data, each with their own personality and 
purpose. You can check what type your experiment is using `get_exp_type()`.

```{r setup}
library(glyexp)
library(glyrepr)
```

## What's the big deal with experiment types?

When we talk about experiment types, we're really talking about two things:

1. A label stored in the `exp_type` field of your `experiment()` object's metadata
2. The expected structure of your variable information table

Here's the thing: when your experiment has a specific type, all the functions 
in glycoverse know what to expect from your data structure. It's like having 
a contract between your data and the analysis functions. Don't worry though—
these are more like friendly guidelines than iron-clad rules. We'll show you 
what this means in practice with some examples.

**Quick tip:** If you're not diving into `glydet` or `glymotif`, you can 
safely skip the `traitomics` and `traitproteomics` sections and focus on 
the first two types.

## Meet the four experiment types

The `glyexp` package recognizes four distinct experiment types:

- **"glycomics"** - Your classic glycan-focused experiments
- **"glycoproteomics"** - When glycans meet proteins  
- **"traitomics"** - Derived traits from glycan data
- **"traitproteomics"** - Derived traits at the protein level

The first two are your bread-and-butter glycobiology experiments. When you 
use `glyread` to import your data, it'll automatically figure out whether 
you're dealing with glycomics or glycoproteomics and set the type accordingly—
pretty neat, right?

The last two are special creatures unique to the glycoverse ecosystem. They're 
what you get when functions from `glymotif` or `glydet` work their magic on 
your data. Let's dive into each type to see what makes them tick.

### Glycomics

Let's start with the most straightforward type: "glycomics". This is your 
classic glycan-focused experiment where each row in the variable information 
table represents a single glycan (or a spectrum match if you're working with MS data).

What to expect in your variable information table:

- `glycan_composition`: The glycan's composition as a `glyrepr::glycan_composition()` object
- `glycan_structure`: (Optional) The glycan's structure as a `glyrepr::glycan_structure()` object

Here's what this looks like in practice:

```{r}
get_var_info(real_experiment2)
```

### Glycoproteomics

Now we step up to "glycoproteomics"—where things get more interesting! Here, 
each row represents a glycopeptide or glycoform (or their peptide spectrum 
matches). Think of it as glycomics data with protein context, telling you 
not just what glycans are there, but where they're attached.

Your variable information table should include:

- `protein`: The protein UniProt accession (character string)
- `protein_site`: The glycosylation site position on the protein (integer)
- `glycan_composition`: The glycan composition as a `glyrepr::glycan_composition()` object
- `glycan_structure`: (Optional) The glycan structure as a `glyrepr::glycan_structure()` object

Here's a real example to illustrate:

```{r}
get_var_info(real_experiment)
```

### Traitomics

Here's where things get uniquely glycoverse! Both `traitomics` and 
`traitproteomics` are special experiment types that emerge when you 
transform your data using `glydet::derive_traits()` or 
`glymotif::quantify_motifs()`.

Starting with `traitomics`: when you feed a `glycomics` experiment to 
`derive_traits()`, you get back a `traitomics` experiment. Instead of 
measuring individual glycans in each sample, you're now looking at 
derived traits—think of it as a higher-level summary of your glycan data.

The variable information table is pretty flexible here—no mandatory columns. 
However, you'll typically see:

- A `trait` column (if using `derive_traits()`)
- A `motif` column (if using `quantify_motifs()`)

### Traitproteomics

Last but not least, we have `traitproteomics`—the most sophisticated of our 
experiment types. When you apply `derive_traits()` to a `glycoproteomics` 
experiment, you get a `traitproteomics` experiment back. 

Here's the key insight: instead of tracking individual glycans at each 
protein site, you're now measuring derived traits at each site. Think of 
each glycosylation site as having its own little glycome, and you're 
summarizing the characteristics of that mini-glycome.

Essential columns in the variable information table:

- `protein`: The protein UniProt accession (character string)  
- `protein_site`: The glycosylation site position (integer)

Plus you'll get:

- A `trait` column (from `derive_traits()`)
- A `motif` column (from `quantify_motifs()`)

Here's a key difference: in `glycoproteomics`, different sites can have 
completely different glycan repertoires. But in `traitproteomics`, every 
site gets scored on the same set of traits or motifs—it's like having 
standardized metrics across all your glycosylation sites.

## Do you need to worry about experiment types?

The short answer? Not really! Experiment types work behind the scenes quite 
intuitively. You'll mostly encounter them in two situations:

1. **Reading documentation** - When you see references to specific experiment types
2. **Error messages** - Like "Expecting a `glycoproteomics` experiment, but got a `traitproteomics` experiment"

The reason these types matter is that different functions have different expectations.
For instance:

- `glymotif::quantify_motifs()` wants `glycomics` or `glycoproteomics` data  
- `glystats::gly_enrich_go()` expects `glycoproteomics` or `traitproteomics` experiments

**When you might need to care:** The main scenario is when you're manually 
creating an experiment object using `experiment()`. In that case, you'll 
want to set the `exp_type` field to match your data structure.

**Important warning:** Once you've created an experiment object, resist the 
temptation to manually fiddle with the `exp_type` field. This will break 
the implicit contract between your data and the glycoverse functions—and 
nobody wants that!