---
title: "Get Started with glyexp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with glyexp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this context, you typically work with three types of data in glycomics and glycoproteomics experiments:

1. **Expression data (assay)** - the actual measurements of your biological molecules (glycans, glycopeptides, etc.)
2. **Molecular annotations (row data)** - the identifiers for your molecules (structures, sequences, etc.)
3. **Experimental metadata (column data)** - the context of your samples (time points, treatments, experimental conditions)

The `SummarizedExperiment` class in the `SummarizedExperiment` packag
 serves as a structured container that keeps all three data types organized and interconnected.
 `glyexp` provides two subclasses of `SummarizedExperiment`:
 `GlycomicSE` for glycomics data, and `GlycoproteomicSE` for glycoproteomics data.

```{r setup, include=FALSE}
library(glyexp)
library(SummarizedExperiment)
library(dplyr)
library(conflicted)

# Resolve function conflicts - prefer glyexp version over deprecated dplyr version
# `dplyr::select_row` is deprecated anyway, so we can safely override it
conflicts_prefer(glyexp::select_row)
```

## For those familiar with SummarizedExperiment

`GlycomicSE` and `GlycoproteomicSE` are both subclasses of `SummarizedExperiment` with a few restrictions:

1. `GlycomicSE` requires a `glycan_composition` column as a `glyrepr::glycan_composition()` vector in the `rowData`.
2. `GlycoproteomicSE` requires `glycan_composition`, as well as `protein` (character) and `protein_site` (integer) columns.
3. There can be only one assay.

That's all.
All your familiar operations provided by `SummarizedExperiment` and `tidySummarizedExperiment`
can also be used on `GlycomicSE` and `GlycoproteomicSE`.

## Getting Started with glyexp

Let's begin with a simple example to illustrate the basic concepts.

```{r}
# `real_experiment` is a built-in `GlycoproteomicSE` object
exp <- real_experiment
exp
```

The summary provides an overview of your entire experiment - variables, observations, and all the metadata.

### The Expression Matrix

The expression matrix contains your numerical data - rows are variables (molecules), columns are observations (samples).

```{r}
assay(exp)[1:5, 1:3]
```

### Variable Information

The variable information table (or "raw data", in the `SummarizedExperiment` terminology)
contains detailed annotations for each molecule.

```{r}
head(rowData(exp))
```

### Sample Information

The sample information table (or "column data") records the experimental conditions for each sample.

```{r}
head(colData(exp))
```

### Rownames and colnames

Row names in `rowData()` and `colData()` match the row and column names in the assay, repectively.

```{r}
identical(rownames(rowData(exp)), rownames(assay(exp)))
identical(rownames(colData(exp)), colnames(assay(exp)))
```

As a shorthand, you can just use `rownames()` and `colnames()` on the container.

```{r}
rownames(exp)[1:5]
colnames(exp)[1:5]
```

## Data Manipulation with glyexp

glyexp provides dplyr-style functions for manipulating `SummarizedExperiment` objects.

For every dplyr function, glyexp provides two specialized versions:

- **`_col()`** functions: work on your sample metadata
- **`_row()`** functions: work on your variable annotations

Here's an example of filtering for group "A" samples:

```{r}
subset_exp <- filter_col(exp, group == "H")
```

Let's check what happened to our sample info:

```{r}
colData(subset_exp)
```

Check the expression matrix:

```{r}
assay(subset_exp)[1:5,]
```

The expression matrix is automatically filtered to match!

This is `filter_col()`: it filters the sample information and automatically updates the expression matrix to match.

Variable filtering works the same way:

```{r}
exp |> filter_col(group == "H")
```

Notice how these functions support the pipe operator (`|>`)? 
That's the `dplyr` DNA in action!

The pattern is straightforward: glyexp functions expect and return
`SummarizedExperiment` objects, including `GlycomicSE` and `GlycoproteomicSE`,
and they preserve the row and column identifiers during operations.

To learn more about `dplyr`-style operations, check out
[this vignette](https://glycoverse.github.io/glyexp/articles/dplyr-style-functions.html).
To learn basic `SummarizedExperiment` operations, check out
[this vignette](https://bioconductor.org/packages//release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html).

## Background and Design Principles

glyexp builds on insights from related data containers:

**SummarizedExperiment**
The foundational omics data container from Bioconductor. Well-established for RNA-seq analysis.

**tidySummarizedExperiment**
An attempt to bring tidy principles to SummarizedExperiment from the tidySummarizedExperiment package.
While the concept is sound, storing all components in a single tibble doesn't align
with the mental model of separated data types.

**massdataset**
A related package for mass spectrometry data.
It provides tidy operations, clean data separation, and data processing history tracking.
We appreciate its approach to reproducibility.

While object-oriented programming has its merits, glyexp takes a functional programming approach.
Your analysis code serves as the reproducibility record - clear, transparent, and familiar to R users.

**Design Philosophy**
glyexp uses functional programming because it aligns with how R users work.
The design emphasizes clear, chainable functions.

Thank you to all the developers who contributed to these foundational packages.

## What's Next?

Now you have the basic understanding of `glyexp`.
Next, you can learn how to use it in your analysis.

- [Creating Experiments](https://glycoverse.github.io/glyexp/articles/create-exp.html)
- [Dplyr-Style Functions](https://glycoverse.github.io/glyexp/articles/dplyr-style-functions.html)
